---
title: GPCA at Scale and Special Cases
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: teal
  preset: homage
resource_files:
- albers.css
- albers.js
- albers-header.html

vignette: >
  %\VignetteIndexEntry{GPCA at Scale and Special Cases}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
options(rmarkdown.html_vignette.check_title = FALSE)
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  message    = FALSE,
  warning    = FALSE,
  fig.width  = 6,
  fig.height = 4,
  out.width  = "85%"
)
library(genpca)
library(Matrix)
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```



This vignette walks through the choices that matter once your data outgrows the defaults: which backend to pick, when to switch to a covariance-only fit, and how to project out-of-sample observations.

## Backend selection

| Method | Best for | Pros | Cons |
|:--|:--|:--|:--|
| `eigen` | Small / medium dense problems | Robust reference behaviour | Can be expensive at scale; very large sparse constraints may trigger truncated eigensolve via `maxeig` |
| `spectra` | Larger matrix-free solves | Lower memory than dense eigendecomp | Iterative behaviour can vary by conditioning |
| `randomized` | Wide (`p >> n`) low-rank workloads | Fast block GEMM / SpMM path | Approximation error depends on tuning |
| `deflation` | Few components, tight memory | Low memory footprint | Can converge slowly; monitor iteration warnings |
| `auto` | Default production usage | Picks among the dense and randomized paths | Heuristics may not be optimal for every regime |

In practice, leave it on `"auto"` unless you have a reason to pin a backend.

## Backends on the same problem

A small head-to-head on a dense problem so you can see how the singular values agree across paths:

```{r backend-bench}
set.seed(11)
n <- 150; p <- 60
X <- matrix(rnorm(n * p), n, p)

t_eig <- system.time(
  fit_eig <- genpca(X, ncomp = 8, method = "eigen",
                    preproc = multivarious::center())
)
t_rnd <- system.time(
  fit_rnd <- genpca(X, ncomp = 8, method = "randomized",
                    preproc = multivarious::center())
)
data.frame(method = c("eigen", "randomized"),
           elapsed = c(t_eig["elapsed"], t_rnd["elapsed"]),
           top_sv  = c(fit_eig$sdev[1], fit_rnd$sdev[1]))
```

```{r backend-plot, echo = FALSE, fig.cap = "Singular values from the eigen and randomized paths agree to plotting precision on this dense problem.", fig.height = 3.5}
plot(fit_eig$sdev, type = "b", pch = 19, col = "grey30",
     ylim = range(c(fit_eig$sdev, fit_rnd$sdev)),
     xlab = "Component", ylab = "Singular value",
     main = "Backend comparison")
lines(fit_rnd$sdev, type = "b", pch = 21, col = "steelblue")
legend("topright", legend = c("eigen", "randomized"),
       col = c("grey30", "steelblue"), pch = c(19, 21),
       bty = "n", cex = 0.85)
```

## Sparse workflow (`spectra`)

The `spectra` backend uses an iterative C++ solver and is the right choice for large sparse problems. The chunk below is shown but not evaluated to keep the vignette fast; it is the pattern to copy:

```{r spectra, eval = FALSE}
set.seed(42)
X_sparse <- rsparsematrix(800, 300, density = 0.005)
fit_sp <- genpca(X_sparse, ncomp = 5, method = "spectra",
                 preproc = multivarious::pass(),
                 constraints_remedy = "ridge")
fit_sp$sdev
```

## Covariance-only GPCA

When you already have the cross-product `C = X' M X`, `genpca_cov()` avoids touching the full data matrix:

```{r cov}
set.seed(123)
n <- 100; p <- 15
X <- matrix(rnorm(n * p), n, p)
M <- diag(runif(n, 0.8, 1.2))
A <- diag(runif(p, 0.7, 1.3))
C <- t(X) %*% M %*% X
fit_cov <- genpca_cov(C, R = A, ncomp = 5, method = "gmd")
fit_cov$d
```

```{r cov-plot, echo = FALSE, fig.cap = "Singular values from the covariance-only fit.", fig.height = 3}
barplot(fit_cov$d, names.arg = paste0("PC", seq_along(fit_cov$d)),
        col = "grey60", border = NA, ylab = "Singular value")
```

## Out-of-sample projection

Fit on training rows, then project held-out observations into the same component space:

```{r oos}
set.seed(7)
X <- matrix(rnorm(200 * 30), 200, 30)
fit <- genpca(X[1:150, ], ncomp = 4,
              preproc = multivarious::center())
scores_test <- multivarious::project(fit, X[151:200, ])
head(scores_test, 4)
```

```{r oos-plot, echo = FALSE, fig.cap = "Training scores (grey) and out-of-sample scores (blue) projected into the same component space.", fig.height = 4}
S_train <- multivarious::scores(fit)
plot(rbind(S_train, scores_test)[, 1:2], type = "n",
     xlab = "PC1", ylab = "PC2",
     main = "Training vs out-of-sample")
points(S_train[, 1], S_train[, 2], pch = 19, col = "grey60")
points(scores_test[, 1], scores_test[, 2], pch = 19, col = "steelblue")
legend("topright", legend = c("Train", "OOS"),
       col = c("grey60", "steelblue"), pch = 19,
       bty = "n", cex = 0.85)
```

## Performance tips

Keep data sparse when possible; avoid centering dense copies if you plan to use `spectra`. Use `constraints_remedy = "ridge"` for empirical metrics. For large problems, limit `ncomp` to what you truly need, and consider the covariance route when `n` is huge but `p` is moderate.

## Where next

See `vignette("gpca-metrics")` for building metrics, and `vignette("genpca")` for a getting-started walkthrough.
