---
title: 'GPCA Metrics: Building M and A'
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: interaction
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts

vignette: |
  %\VignetteIndexEntry{GPCA Metrics: Building M and A}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
if (requireNamespace("ragg", quietly = TRUE)) knitr::opts_chunk$set(dev = "ragg_png")
if (requireNamespace("systemfonts", quietly = TRUE)) albersdown::albers_register_fonts()
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
options(rmarkdown.html_vignette.check_title = FALSE)
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  message    = FALSE,
  warning    = FALSE,
  fig.width  = 6,
  fig.height = 4,
  out.width  = "85%"
)
library(genpca)
library(Matrix)

flip <- function(M) as.matrix(M)[, ncol(M):1]
heat <- function(M, main = "") {
  image(flip(M), axes = FALSE, main = main,
        col = grey.colors(20, start = 0.95, end = 0.2))
}
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```




This vignette collects practical recipes for row and column metrics, plus notes on SPD remedies and the experimental `gpca_mle()` learner.

## Why metrics

Metrics encode weighting and correlation. The row metric `M` changes how observations are compared. The column metric `A` changes how variables are compared. Both must be symmetric positive **semi-definite** (PSD) for GPCA -- singular metrics are accepted (e.g. a rank-deficient graph Laplacian), and `constraints_remedy` can repair mildly invalid input. In return, you can bake known structure -- temporal correlation, spatial proximity, group membership, heteroscedastic noise -- straight into the decomposition.

Structure alone is not enough, though: you also have to get the *direction*
right, and supplying a matrix versus its inverse produces opposite results.
The next section is about that, and it is worth reading before the recipes.

## Heteroscedastic diagonals

The simplest non-trivial metric is a diagonal that down-weights noisy rows or columns:

```{r hetero-demo}
set.seed(42)
n <- 60; p <- 20
X <- matrix(rnorm(n * p), n, p)

col_noise_sd <- runif(p, 0.5, 2)
A <- Diagonal(x = 1 / col_noise_sd^2)
row_noise_sd <- runif(n, 0.7, 1.3)
M <- Diagonal(x = 1 / row_noise_sd^2)

fit <- genpca(X, M = M, A = A, ncomp = 3,
              preproc = multivarious::center())
fit$sdev
```

```{r hetero-plot, echo = FALSE, fig.cap = "Inverse-variance weights on columns (top) and rows (bottom). Noisier dimensions get smaller weights.", fig.height = 3.5}
op <- par(mfrow = c(2, 1), mar = c(2.5, 4, 2, 1))
barplot(diag(A), border = NA, col = "steelblue",
        main = "Column weights diag(A)", names.arg = NA)
barplot(diag(M), border = NA, col = "tomato",
        main = "Row weights diag(M)", names.arg = NA)
par(op)
```

## Which way does a metric point?

Before the recipes, the single most important thing to get right — and the
easiest to get backwards. **A metric amplifies its own dominant
eigendirections.** Loadings are `components(fit)` $= A\,$`ov`, so whichever
patterns $A$ assigns large eigenvalues to are the patterns that come out of
the decomposition.

For a spatial or temporal structure there are two natural matrices, and they
point in *opposite* directions:

| You supply | $v^{\top}Av$ measures | Large eigenvalues on | Components come out |
|:--|:--|:--|:--|
| A **smoother**: kernel $K$, adjacency $I + \alpha W$, $(I+\alpha L)^{-1}$, heat kernel $e^{-tL}$ | agreement between neighbours | smooth patterns | **smoother** |
| A **precision**: Laplacian $L$, $I + \alpha L$, $K^{-1}$, inverse AR(1) | disagreement across edges (Dirichlet energy $\sum_{i\sim j}(v_i - v_j)^2$) | rough patterns | **rougher** |

Both are legitimate, because they encode different beliefs about *where the
noise lives*. Recall that $A = \Sigma_{\text{col}}^{-1}$:

- **Smoother as metric** $\Rightarrow$ $\Sigma_{\text{col}}$ has its variance in
  the rough directions $\Rightarrow$ "the noise is high-frequency speckle; the
  signal is smooth." This is denoising, and it is what you want when you are
  after spatially coherent maps.
- **Precision as metric** $\Rightarrow$ $\Sigma_{\text{col}}$ has its variance in
  the smooth directions $\Rightarrow$ "a smooth field — drift, a scanner
  gradient, a global trend — is the nuisance." Whitening it away lets
  fine-scale structure surface. This is ordinary generalized least squares
  against correlated noise.

So the question is never "adjacency or Laplacian?" in the abstract. It is:
**is the smooth thing my signal, or my nuisance?**

```{r orientation-demo, echo = FALSE, fig.cap = "The same graph, two metrics. Left: a smoother concentrates PC1 on the smooth blob. Right: the Laplacian whitens the smooth field away, so PC1 locks onto the fine-scale checkerboard that was buried underneath it.", fig.width = 7, fig.height = 3.5}
set.seed(7)
gg <- 12; pp <- gg * gg; nn <- 80
gidx <- expand.grid(r = 1:gg, c = 1:gg)
Wg <- matrix(0, pp, pp)
for (i in 1:pp) for (j in 1:pp)
  if (i < j && abs(gidx$r[i] - gidx$r[j]) + abs(gidx$c[i] - gidx$c[j]) == 1) {
    Wg[i, j] <- 1; Wg[j, i] <- 1
  }
Lg <- diag(rowSums(Wg)) - Wg
blob <- exp(-((gidx$r - 4)^2 + (gidx$c - 4)^2) / 6); blob <- blob / sqrt(sum(blob^2))
chk  <- as.numeric(((gidx$r + gidx$c) %% 2) * 2 - 1); chk <- chk / sqrt(sum(chk^2))
Xg <- matrix(rnorm(nn), nn, 1) %*% t(chk) * 3 +
      matrix(rnorm(nn), nn, 1) %*% t(blob) * 15 +
      matrix(rnorm(nn * pp), nn, pp) * 0.5
f_sm  <- genpca(Xg, A = solve(diag(pp) + 2 * Lg), ncomp = 1, preproc = multivarious::center())
f_lap <- genpca(Xg, A = diag(pp) + 50 * Lg,      ncomp = 1, preproc = multivarious::center())
op <- par(mfrow = c(1, 2), mar = c(1, 1, 3, 1))
for (ff in list(list(f_sm, "Smoother metric: finds the blob"),
                list(f_lap, "Laplacian metric: finds the checkerboard"))) {
  v <- multivarious::components(ff[[1]])[, 1]
  image(matrix(v, gg, gg), axes = FALSE, main = ff[[2]], cex.main = 0.95,
        col = grey.colors(24, start = 0.95, end = 0.15))
}
par(op)
```

Two practical notes on the graph matrices themselves. A raw adjacency $W$ is
**indefinite** (its eigenvalues sum to zero), so it is not a valid metric on
its own — shift it, as in $I + \alpha W$ with $\alpha$ small enough to keep
it PSD. A raw Laplacian $L$ is PSD but **singular**: $L\mathbf{1} = 0$, so
the spatially constant pattern has zero length under it. `genpca()` accepts
that (the whitening uses a pseudo-inverse), but adding a small ridge,
$L + \varepsilon I$, is usually what you want.

Finally, the strength matters and is not cosmetic. In the example above the
first component only switched from the smooth blob to the checkerboard once
$\alpha$ passed roughly 50; at $\alpha = 2$ the blob still dominated. Sweep
it and look at your loadings.

## Recipes

Each recipe below is labelled with the direction it produces. All three are
written here as **precision** matrices, i.e. the noise-whitening stance; drop
the `solve()` to get the smoothing version instead.

### AR(1) row metric (whitens temporal autocorrelation)

Standard GLS treatment of serially correlated observations, as in fMRI
prewhitening: it removes temporal autocorrelation rather than imposing
temporal smoothness.

```{r ar1}
rho     <- 0.7
n_t     <- 60
idx     <- 0:(n_t - 1)
Sigma_r <- outer(idx, idx, function(i, j) rho^abs(i - j))
M_ar1   <- solve(Sigma_r + 1e-3 * diag(n_t))
```

### Spatial RBF kernel (whitens smooth spatial noise)

```{r rbf}
coords <- as.matrix(expand.grid(x = 1:8, y = 1:8))
d2     <- as.matrix(dist(coords))^2
ell    <- 2
K      <- exp(-d2 / (2 * ell^2))
A_rbf  <- solve(K + 1e-3 * diag(nrow(K)))   # precision: emphasises fine scale
# A_smooth <- K + 1e-3 * diag(nrow(K))      # kernel itself: smooth loadings
```

### Graph Laplacian (emphasises contrast across edges)

```{r lap}
W <- bandSparse(30, k = c(-1, 0, 1),
                diagonals = list(rep(0.2, 29),
                                 rep(1, 30),
                                 rep(0.2, 29)))
D     <- Diagonal(x = rowSums(W))
A_lap <- (D - W) + 1e-2 * Diagonal(nrow(W))
# A_smooth <- solve(A_lap)                  # smoother: spatially coherent loadings
```

```{r recipe-plots, echo = FALSE, fig.cap = "Three structured metrics, all shown in the precision (noise-whitening) orientation. Off-diagonal banding is what couples nearby rows or variables -- it tells GPCA 'treat these dimensions as related, not independent'.", fig.width = 8, fig.height = 3}
op <- par(mfrow = c(1, 3), mar = c(2, 2, 2, 1))
heat(M_ar1, "Inverse AR(1)")
heat(A_rbf, "Inverse RBF kernel")
heat(A_lap, "Regularised Laplacian")
par(op)
```

### A note on `sfpca()`

`sfpca()` takes the *opposite* input for the same intent. There the structure
enters as a constraint, $v^{\top}(I + \alpha\Omega)v \le 1$, which charges
rough $v$ against a fixed budget — so you pass the roughness operator
(Laplacian, second differences) directly via `alpha_v`, and larger `alpha_v`
means **smoother**. Metric form and constraint form are inverse to one
another: the same Laplacian smooths in `sfpca()` and roughens in `genpca()`.

## Learning metrics with `gpca_mle()`

When you suspect structured noise but lack good priors, `gpca_mle()` learns SPD metrics by alternating GPCA with matrix-normal maximum likelihood:

```{r mle-fit}
set.seed(1)
n_m <- 40; p_m <- 10
X_mle <- matrix(rnorm(n_m * p_m), n_m, p_m)
fit_mle <- gpca_mle(X_mle, ncomp = 2, max_iter = 6,
                    lambda = 1e-3, scale_fix = "trace",
                    method = "eigen", verbose = FALSE)
range(diag(as.matrix(fit_mle$M)))
range(diag(as.matrix(fit_mle$A)))
```

```{r mle-plot, echo = FALSE, fig.cap = "Learned row metric M (left) and column metric A (right) on i.i.d. Gaussian data. With identity-noise input the learner stays close to identity, with mild row- and column-specific damping."}
op <- par(mfrow = c(1, 2), mar = c(2, 2, 2, 1))
heat(fit_mle$M, "Learned M")
heat(fit_mle$A, "Learned A")
par(op)
```

Keep `lambda` non-zero, start with a small `ncomp`, and watch warnings -- they signal a metric was repaired during the iteration.

## SPD remedies in practice

- `constraints_remedy = "ridge"` (default) adds a diagonal jitter to keep metrics SPD.
- `"clip"` truncates tiny negative eigenvalues; `"identity"` falls back to identity if a metric misbehaves.
- Scale metrics before use (e.g., divide by mean diagonal) to avoid ill-conditioning.
- After fitting, check `range(eigen(A)$values)` on small problems to verify conditioning.

## Where next

See `vignette("structured-noise")` for how to choose the transfer function when several kinds of structure are present at once, and `vignette("gpca-scale")` for backend choices, sparse workflows, and covariance-only GPCA.
