---
title: 'GPCA Metrics: Building M and A'
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: teal
  preset: homage
resource_files:
- albers.css
- albers.js
- albers-header.html

vignette: |
  %\VignetteIndexEntry{GPCA Metrics: Building M and A} %\VignetteEngine{knitr::rmarkdown} %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
options(rmarkdown.html_vignette.check_title = FALSE)
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  message    = FALSE,
  warning    = FALSE,
  fig.width  = 6,
  fig.height = 4,
  out.width  = "85%"
)
library(genpca)
library(Matrix)

flip <- function(M) as.matrix(M)[, ncol(M):1]
heat <- function(M, main = "") {
  image(flip(M), axes = FALSE, main = main,
        col = grey.colors(20, start = 0.95, end = 0.2))
}
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```



This vignette collects practical recipes for row and column metrics, plus notes on SPD remedies and the experimental `gpca_mle()` learner.

## Why metrics

Metrics encode weighting and correlation. The row metric `M` changes how observations are compared. The column metric `A` changes how variables are compared. Both must be symmetric positive definite (SPD) for GPCA. In return, you can bake known structure -- temporal smoothness, spatial proximity, group membership, heteroscedastic noise -- straight into the decomposition.

## Heteroscedastic diagonals

The simplest non-trivial metric is a diagonal that down-weights noisy rows or columns:

```{r hetero-demo}
set.seed(42)
n <- 60; p <- 20
X <- matrix(rnorm(n * p), n, p)

col_noise_sd <- runif(p, 0.5, 2)
A <- Diagonal(x = 1 / col_noise_sd^2)
row_noise_sd <- runif(n, 0.7, 1.3)
M <- Diagonal(x = 1 / row_noise_sd^2)

fit <- genpca(X, M = M, A = A, ncomp = 3,
              preproc = multivarious::center())
fit$sdev
```

```{r hetero-plot, echo = FALSE, fig.cap = "Inverse-variance weights on columns (top) and rows (bottom). Noisier dimensions get smaller weights.", fig.height = 3.5}
op <- par(mfrow = c(2, 1), mar = c(2.5, 4, 2, 1))
barplot(diag(A), border = NA, col = "steelblue",
        main = "Column weights diag(A)", names.arg = NA)
barplot(diag(M), border = NA, col = "tomato",
        main = "Row weights diag(M)", names.arg = NA)
par(op)
```

## Recipes

These three patterns -- AR(1) smoothing, spatial RBF kernel, graph Laplacian -- cover most structured-noise applications.

### AR(1) row metric (smoothness in time)

```{r ar1}
rho     <- 0.7
n_t     <- 60
idx     <- 0:(n_t - 1)
Sigma_r <- outer(idx, idx, function(i, j) rho^abs(i - j))
M_ar1   <- solve(Sigma_r + 1e-3 * diag(n_t))
```

### Spatial RBF kernel

```{r rbf}
coords <- as.matrix(expand.grid(x = 1:8, y = 1:8))
d2     <- as.matrix(dist(coords))^2
ell    <- 2
K      <- exp(-d2 / (2 * ell^2))
A_rbf  <- solve(K + 1e-3 * diag(nrow(K)))
```

### Graph Laplacian

```{r lap}
W <- bandSparse(30, k = c(-1, 0, 1),
                diagonals = list(rep(0.2, 29),
                                 rep(1, 30),
                                 rep(0.2, 29)))
D     <- Diagonal(x = rowSums(W))
A_lap <- (D - W) + 1e-2 * Diagonal(nrow(W))
```

```{r recipe-plots, echo = FALSE, fig.cap = "Three structured metrics. Off-diagonal banding is what couples nearby rows or variables -- it tells GPCA 'treat these dimensions as related, not independent'.", fig.width = 8, fig.height = 3}
op <- par(mfrow = c(1, 3), mar = c(2, 2, 2, 1))
heat(M_ar1, "Inverse AR(1)")
heat(A_rbf, "Inverse RBF kernel")
heat(A_lap, "Regularised Laplacian")
par(op)
```

## Learning metrics with `gpca_mle()`

When you suspect structured noise but lack good priors, `gpca_mle()` learns SPD metrics by alternating GPCA with matrix-normal maximum likelihood:

```{r mle-fit}
set.seed(1)
n_m <- 40; p_m <- 10
X_mle <- matrix(rnorm(n_m * p_m), n_m, p_m)
fit_mle <- gpca_mle(X_mle, ncomp = 2, max_iter = 6,
                    lambda = 1e-3, scale_fix = "trace",
                    method = "eigen", verbose = FALSE)
range(diag(as.matrix(fit_mle$M)))
range(diag(as.matrix(fit_mle$A)))
```

```{r mle-plot, echo = FALSE, fig.cap = "Learned row metric M (left) and column metric A (right) on i.i.d. Gaussian data. With identity-noise input the learner stays close to identity, with mild row- and column-specific damping."}
op <- par(mfrow = c(1, 2), mar = c(2, 2, 2, 1))
heat(fit_mle$M, "Learned M")
heat(fit_mle$A, "Learned A")
par(op)
```

Keep `lambda` non-zero, start with a small `ncomp`, and watch warnings -- they signal a metric was repaired during the iteration.

## SPD remedies in practice

- `constraints_remedy = "ridge"` (default) adds a diagonal jitter to keep metrics SPD.
- `"clip"` truncates tiny negative eigenvalues; `"identity"` falls back to identity if a metric misbehaves.
- Scale metrics before use (e.g., divide by mean diagonal) to avoid ill-conditioning.
- After fitting, check `range(eigen(A)$values)` on small problems to verify conditioning.

## Where next

See `vignette("gpca-scale")` for backend choices, sparse workflows, and covariance-only GPCA.
