---
title: 'GPCA Metrics: Building M and A'
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: interaction
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts

vignette: |
  %\VignetteIndexEntry{GPCA Metrics: Building M and A}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
if (requireNamespace("ragg", quietly = TRUE)) knitr::opts_chunk$set(dev = "ragg_png")
if (requireNamespace("systemfonts", quietly = TRUE)) albersdown::albers_register_fonts()
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
options(rmarkdown.html_vignette.check_title = FALSE)
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  message    = FALSE,
  warning    = FALSE,
  fig.width  = 6,
  fig.height = 4,
  out.width  = "85%"
)
library(genpca)
library(Matrix)

flip <- function(M) as.matrix(M)[, ncol(M):1]
heat <- function(M, main = "") {
  image(flip(M), axes = FALSE, main = main,
        col = grey.colors(20, start = 0.95, end = 0.2))
}
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```




This vignette collects practical recipes for row and column metrics, plus notes on SPD remedies and the experimental `gpca_mle()` learner.

## Why metrics

Metrics encode weighting and correlation. The row metric `M` changes how
observations are compared; the column metric `A` changes how variables are
compared. Setting either to something other than the identity is how you tell
the decomposition what you already know about the data: that the samples are a
time series, that the variables sit on a spatial grid or fall into groups, that
some measurements are noisier than others. Ordinary PCA has no way to accept
that information and treats every row and column alike. GPCA writes it into the
objective being optimised.

Structure alone is not enough, though: you also have to get the *direction*
right, and supplying a matrix versus its inverse produces opposite results.
The next section is about that, and it is worth reading before the recipes.

Both metrics must also be symmetric and positive semi-definite. That rarely
gets in the way, but it does constrain what you can pass; see
[SPD requirements and remedies](#spd-requirements-and-remedies) below for what
the requirement means and what `genpca()` does when a metric falls short of it.

## Heteroscedastic diagonals

The simplest non-trivial metric is a diagonal that down-weights noisy rows or columns:

```{r hetero-demo}
set.seed(42)
n <- 60; p <- 20
X <- matrix(rnorm(n * p), n, p)

col_noise_sd <- runif(p, 0.5, 2)
A <- Diagonal(x = 1 / col_noise_sd^2)
row_noise_sd <- runif(n, 0.7, 1.3)
M <- Diagonal(x = 1 / row_noise_sd^2)

fit <- genpca(X, M = M, A = A, ncomp = 3,
              preproc = multivarious::center())
fit$sdev
```

```{r hetero-plot, echo = FALSE, fig.cap = "Inverse-variance weights on columns (top) and rows (bottom). Noisier dimensions get smaller weights.", fig.height = 3.5}
op <- par(mfrow = c(2, 1), mar = c(2.5, 4, 2, 1))
barplot(diag(A), border = NA, col = "steelblue",
        main = "Column weights diag(A)", names.arg = NA)
barplot(diag(M), border = NA, col = "tomato",
        main = "Row weights diag(M)", names.arg = NA)
par(op)
```

### Inverse column variance *is* scaled PCA

Inverse-variance weighting on the columns is not a new idea in disguise: it is
exactly the standardisation that `prcomp(scale. = TRUE)` performs. GPCA whitens
with the square root $A^{1/2}$, so setting $A = \operatorname{diag}(1/s_j^2)$
makes $X A^{1/2} = X \operatorname{diag}(1/s_j)$ — the column-scaled matrix
that correlation-matrix PCA decomposes.

```{r scaled-pca-equiv}
set.seed(42)
Xv <- matrix(rnorm(60 * 20), 60, 20) %*% diag(runif(20, 0.5, 3))
sds <- apply(Xv, 2, sd)

g  <- genpca(Xv, A = Diagonal(x = 1 / sds^2), ncomp = 5,
             preproc = multivarious::center())
pr <- prcomp(Xv, scale. = TRUE)

# scores agree component by component
sapply(1:3, function(k) cor(multivarious::scores(g)[, k], pr$x[, k]))
```

The scores are identical (up to sign). The reported `sdev` values differ by a
single constant, because `prcomp()` divides its singular values by
$\sqrt{n-1}$ and `genpca()` reports them unnormalised:

```{r scaled-pca-sdev}
rbind(genpca = g$sdev[1:5],
      prcomp = pr$sdev[1:5],
      ratio  = g$sdev[1:5] / pr$sdev[1:5])
sqrt(nrow(Xv) - 1)
```

The constant ratio is the whole difference. This is worth internalising as the
baseline: **a diagonal `A` generalises column scaling**, and everything else in
this vignette — kernels, Laplacians, AR(1) structure — generalises it further
by letting the metric go off-diagonal.

### Weighting both margins at once

The example above sets `M` and `A` simultaneously, which is a reasonable thing
to want when both observations and variables are heteroscedastic. Two
consequences are easy to miss.

**The two weightings interact.** GPCA works with $M^{1/2} X A^{1/2}$, so the
row weights change each column's effective variance and the column weights
change each row's. Estimating row and column standard deviations from the raw
data and applying both at once therefore standardises *neither* margin:

```{r both-margins}
Xc <- scale(Xv, center = TRUE, scale = FALSE)
W  <- diag(1 / apply(Xc, 1, sd)) %*% Xc %*% diag(1 / apply(Xc, 2, sd))

range(apply(W, 1, sd))   # row SDs, would be constant if standardised
range(apply(W, 2, sd))   # column SDs
```

It moves in the right direction — the row SDs are markedly more uniform than
in `Xc` — but one pass does not land on unit variance either way, because each
rescaling perturbs the other's normalisation. Getting both margins standardised
requires alternating between them until they settle, which is precisely what
`gpca_mle()` does when it iterates between `M` and `A`.

**Only the product of the two scales is identified.** Replacing $(M, A)$ with
$(cM, A/c)$ leaves $M^{1/2} X A^{1/2}$ untouched, so the fit cannot distinguish
them:

```{r scale-indeterminacy}
M0 <- Diagonal(x = 1 / apply(Xv, 1, sd)^2)
A0 <- Diagonal(x = 1 / sds^2)

f1 <- genpca(Xv, M = M0,     A = A0,     ncomp = 4, preproc = multivarious::center())
f2 <- genpca(Xv, M = 7 * M0, A = A0 / 7, ncomp = 4, preproc = multivarious::center())

max(abs(f1$sdev - f2$sdev))
```

The singular values and the component subspace are identical; only the scores
pick up a constant factor, since their normalisation is tied to the scale of
`M`. The practical upshot is that there is no point tuning the overall
magnitude of `M` against that of `A` — it is the *relative* weighting within
each metric that changes the answer. This indeterminacy is also why
`gpca_mle()` takes a `scale_fix` argument: when both metrics are learned, the
split has to be pinned down by convention rather than by the data.

## Which way does a metric point?

Before the recipes, the single most important thing to get right — and the
easiest to get backwards. **A metric amplifies its own dominant
eigendirections.**

Seeing why means being precise about what comes back from a fit. GPCA
factorises $X \approx U D V^{\top}$, where $V$ is orthonormal *in the column
metric* rather than in the ordinary sense: $V^{\top} A V = I$. The loadings
returned by `components(fit)` are not $V$ but $AV$:

```{r loadings-are-AV}
set.seed(1)
Xd <- matrix(rnorm(400), 40, 10)
Ad <- crossprod(matrix(rnorm(100), 10, 10)) / 10 + diag(10)
fd <- genpca(Xd, A = Ad, ncomp = 3, preproc = multivarious::center())

max(abs(multivarious::components(fd) - as.matrix(Ad %*% fd$ov)))
```

That multiplication by $A$ is the whole story. It stretches every direction in
proportion to the eigenvalue $A$ assigns it, so the patterns $A$ scores highly
are the patterns that dominate the loadings you read off. (The bare factor $V$
is kept in the `ov` slot if you ever need it, but `components()` is what you
should normally interpret.)

For a spatial or temporal structure there are two natural matrices, and they
point in *opposite* directions:

| You supply | $v^{\top}Av$ measures | Large eigenvalues on | Components come out |
|:--|:--|:--|:--|
| A **smoother**: kernel $K$, adjacency $I + \alpha W$, $(I+\alpha L)^{-1}$, heat kernel $e^{-tL}$ | agreement between neighbours | smooth patterns | **smoother** |
| A **precision**: Laplacian $L$, $I + \alpha L$, $K^{-1}$, inverse AR(1) | disagreement across edges (Dirichlet energy $\sum_{i\sim j}(v_i - v_j)^2$) | rough patterns | **rougher** |

Both are legitimate, because they encode different beliefs about *where the
noise lives*. Recall that $A = \Sigma_{\text{col}}^{-1}$:

- **Smoother as metric** $\Rightarrow$ $\Sigma_{\text{col}}$ has its variance in
  the rough directions $\Rightarrow$ "the noise is high-frequency speckle; the
  signal is smooth." This is denoising, and it is what you want when you are
  after spatially coherent maps.
- **Precision as metric** $\Rightarrow$ $\Sigma_{\text{col}}$ has its variance in
  the smooth directions $\Rightarrow$ "a smooth field — drift, a scanner
  gradient, a global trend — is the nuisance." Whitening it away lets
  fine-scale structure surface. This is ordinary generalized least squares
  against correlated noise.

So the question is never "adjacency or Laplacian?" in the abstract. It is:
**is the smooth thing my signal, or my nuisance?**

```{r orientation-demo, echo = FALSE, fig.cap = "The same graph, two metrics. Left: a smoother concentrates PC1 on the smooth blob. Right: the Laplacian whitens the smooth field away, so PC1 locks onto the fine-scale checkerboard that was buried underneath it.", fig.width = 7, fig.height = 3.5}
set.seed(7)
gg <- 12; pp <- gg * gg; nn <- 80
gidx <- expand.grid(r = 1:gg, c = 1:gg)
Wg <- matrix(0, pp, pp)
for (i in 1:pp) for (j in 1:pp)
  if (i < j && abs(gidx$r[i] - gidx$r[j]) + abs(gidx$c[i] - gidx$c[j]) == 1) {
    Wg[i, j] <- 1; Wg[j, i] <- 1
  }
Lg <- diag(rowSums(Wg)) - Wg
blob <- exp(-((gidx$r - 4)^2 + (gidx$c - 4)^2) / 6); blob <- blob / sqrt(sum(blob^2))
chk  <- as.numeric(((gidx$r + gidx$c) %% 2) * 2 - 1); chk <- chk / sqrt(sum(chk^2))
Xg <- matrix(rnorm(nn), nn, 1) %*% t(chk) * 3 +
      matrix(rnorm(nn), nn, 1) %*% t(blob) * 15 +
      matrix(rnorm(nn * pp), nn, pp) * 0.5
f_sm  <- genpca(Xg, A = solve(diag(pp) + 2 * Lg), ncomp = 1, preproc = multivarious::center())
f_lap <- genpca(Xg, A = diag(pp) + 50 * Lg,      ncomp = 1, preproc = multivarious::center())
op <- par(mfrow = c(1, 2), mar = c(1, 1, 3, 1))
for (ff in list(list(f_sm, "Smoother metric: finds the blob"),
                list(f_lap, "Laplacian metric: finds the checkerboard"))) {
  v <- multivarious::components(ff[[1]])[, 1]
  image(matrix(v, gg, gg), axes = FALSE, main = ff[[2]], cex.main = 0.95,
        col = grey.colors(24, start = 0.95, end = 0.15))
}
par(op)
```

Two practical notes on the graph matrices themselves. A raw adjacency $W$ is
**indefinite** (its eigenvalues sum to zero), so it is not a valid metric on
its own — shift it, as in $I + \alpha W$ with $\alpha$ small enough to keep
it PSD. A raw Laplacian $L$ is PSD but **singular**: $L\mathbf{1} = 0$, so
the spatially constant pattern has zero length under it. `genpca()` accepts
that (the whitening uses a pseudo-inverse), but adding a small ridge,
$L + \varepsilon I$, is usually what you want.

Finally, the strength matters and is not cosmetic. In the example above the
first component only switched from the smooth blob to the checkerboard once
$\alpha$ passed roughly 50; at $\alpha = 2$ the blob still dominated. Sweep
it and look at your loadings.

## Recipes

Each recipe below is labelled with the direction it produces. All three are
written here as **precision** matrices, i.e. the noise-whitening stance; drop
the `solve()` to get the smoothing version instead.

### AR(1) row metric (whitens temporal autocorrelation)

Standard GLS treatment of serially correlated observations, as in fMRI
prewhitening: it removes temporal autocorrelation rather than imposing
temporal smoothness.

```{r ar1}
rho     <- 0.7
n_t     <- 60
idx     <- 0:(n_t - 1)
Sigma_r <- outer(idx, idx, function(i, j) rho^abs(i - j))
M_ar1   <- solve(Sigma_r + 1e-3 * diag(n_t))
```

### Spatial RBF kernel (whitens smooth spatial noise)

```{r rbf}
coords <- as.matrix(expand.grid(x = 1:8, y = 1:8))
d2     <- as.matrix(dist(coords))^2
ell    <- 2
K      <- exp(-d2 / (2 * ell^2))
A_rbf  <- solve(K + 1e-3 * diag(nrow(K)))   # precision: emphasises fine scale
# A_smooth <- K + 1e-3 * diag(nrow(K))      # kernel itself: smooth loadings
```

### Graph Laplacian (emphasises contrast across edges)

```{r lap}
W <- bandSparse(30, k = c(-1, 0, 1),
                diagonals = list(rep(0.2, 29),
                                 rep(1, 30),
                                 rep(0.2, 29)))
D     <- Diagonal(x = rowSums(W))
A_lap <- (D - W) + 1e-2 * Diagonal(nrow(W))
# A_smooth <- solve(A_lap)                  # smoother: spatially coherent loadings
```

```{r recipe-plots, echo = FALSE, fig.cap = "Three structured metrics, all shown in the precision (noise-whitening) orientation. Off-diagonal banding is what couples nearby rows or variables -- it tells GPCA 'treat these dimensions as related, not independent'.", fig.width = 8, fig.height = 3}
op <- par(mfrow = c(1, 3), mar = c(2, 2, 2, 1))
heat(M_ar1, "Inverse AR(1)")
heat(A_rbf, "Inverse RBF kernel")
heat(A_lap, "Regularised Laplacian")
par(op)
```

### A note on `sfpca()`

`sfpca()` takes the *opposite* input for the same intent. There the structure
enters as a constraint, $v^{\top}(I + \alpha\Omega)v \le 1$, which charges
rough $v$ against a fixed budget — so you pass the roughness operator
(Laplacian, second differences) directly via `alpha_v`, and larger `alpha_v`
means **smoother**. Metric form and constraint form are inverse to one
another: the same Laplacian smooths in `sfpca()` and roughens in `genpca()`.

## Learning metrics with `gpca_mle()`

When you suspect structured noise but lack good priors, `gpca_mle()` learns SPD metrics by alternating GPCA with matrix-normal maximum likelihood:

```{r mle-fit}
set.seed(1)
n_m <- 40; p_m <- 10
X_mle <- matrix(rnorm(n_m * p_m), n_m, p_m)
fit_mle <- gpca_mle(X_mle, ncomp = 2, max_iter = 6,
                    lambda = 1e-3, scale_fix = "trace",
                    method = "eigen", verbose = FALSE)
range(diag(as.matrix(fit_mle$M)))
range(diag(as.matrix(fit_mle$A)))
```

```{r mle-plot, echo = FALSE, fig.cap = "Learned row metric M (left) and column metric A (right) on i.i.d. Gaussian data. With identity-noise input the learner stays close to identity, with mild row- and column-specific damping."}
op <- par(mfrow = c(1, 2), mar = c(2, 2, 2, 1))
heat(fit_mle$M, "Learned M")
heat(fit_mle$A, "Learned A")
par(op)
```

Keep `lambda` non-zero, start with a small `ncomp`, and watch warnings -- they signal a metric was repaired during the iteration.

## SPD requirements and remedies

### Why the requirement exists

GPCA measures variance in the inner products $\langle u, v\rangle_M = u^\top M
v$ and $\langle x, y\rangle_A = x^\top A y$, and the solvers whiten the data
with the square roots $M^{1/2}$ and $A^{1/2}$. Both steps need the metrics to
be symmetric positive semi-definite (PSD). If a metric has a negative
eigenvalue, vectors in that direction have negative squared length, the square
root is not real, and "maximise variance" no longer picks out anything
meaningful.

Note that *semi*-definite is enough. **Singular metrics are perfectly legal.**
A graph Laplacian is rank-deficient by construction — it has a zero eigenvalue
on the constant vector — and `genpca()` takes it without complaint. A zero
eigenvalue simply means that direction is given no weight. Only *negative*
eigenvalues are a problem.

### What happens when a metric falls short

Metrics usually fail the test for dull reasons rather than modelling mistakes:
a covariance estimated from fewer samples than variables, a kernel matrix with
round-off in the last few digits, a matrix that is asymmetric by $10^{-16}$
because of the order in which it was assembled. The check itself is tolerant —
eigenvalues down to `-tol * max(abs(diag()))` with `tol = 1e-6` count as
non-negative — so ordinary floating-point noise never triggers a remedy, and a
metric that passes is used exactly as supplied.

When a metric does fail, `constraints_remedy` decides what happens next:

| Value | What it does | What it costs |
|---|---|---|
| `"ridge"` (default) | Adds a diagonal shift (from the Gershgorin bound, with a `Matrix::nearPD()` fallback for small dense matrices) just large enough to make the matrix positive definite. Preserves sparsity. | The shift pulls the metric toward a multiple of the identity, diluting the structure you supplied. A *large* shift means the input was badly indefinite — diagnose it rather than absorb it. |
| `"clip"` | Eigendecomposes and sets the negative eigenvalues to zero, leaving the rest of the spectrum exactly as it was. | Densifies the matrix, so it refuses sparse input larger than 2000×2000. Use `"ridge"` at that size. |
| `"identity"` | Replaces the offending metric with the identity. | Discards your structure and quietly hands back ordinary PCA, with the run still reporting success. Pass `verbose = TRUE` to be told when this fires. |
| `"error"` | Refuses the input. | Nothing — this is the right setting when the metric comes from a pipeline that ought to be producing a valid one. `genpca_cov()` defaults to it for that reason. |

Two behaviours are worth knowing about because they are silent:

- `"ridge"` and `"clip"` both symmetrise the input by mirroring the **upper**
  triangle onto the lower. If you pass a genuinely asymmetric matrix, the lower
  triangle is discarded without a word. Only `"error"` reports it.
- `"identity"` and a large `"ridge"` shift both produce a fit that looks
  healthy while encoding much less structure than you intended.

So if a metric matters to the result, verify it rather than trusting the
remedy to have done something sensible:

```{r spd-check, eval = FALSE}
# Is the metric usable as-is? (small problems)
range(eigen(as.matrix(A), symmetric = TRUE, only.values = TRUE)$values)

# Did a remedy fire, and how hard?
fit <- genpca(X, A = A, M = M, ncomp = 3, verbose = TRUE)
```

Two habits that avoid the problem to begin with: scale metrics before use
(dividing by the mean diagonal, say) so that `M` and `A` are comparably
conditioned, and prefer building a metric that is PSD by construction — a
kernel, a Laplacian, an inverse-variance diagonal — over one estimated and then
patched. With `gpca_mle()`, warnings during the iteration mean a metric was
repaired along the way and the learned result should be inspected.

## Where next

See `vignette("structured-noise")` for how to choose the transfer function when several kinds of structure are present at once, and `vignette("gpca-scale")` for backend choices, sparse workflows, and covariance-only GPCA.
