---
title: "GPCA at Scale and Special Cases"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GPCA at Scale and Special Cases}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
options(rmarkdown.html_vignette.check_title = FALSE)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(genpca)
library(Matrix)
```

Backend selection
-----------------

- `method = "auto"`: default; chooses among `eigen`, `spectra`, and `randomized`.
- `method = "eigen"`: small/medium dense problems; reference exact path.
- `method = "spectra"`: large/iterative matrix-free path (requires C++ build).
- `method = "randomized"`: wide low-rank problems with sparse metrics; approximate but often faster.
- `method = "deflation"`: sequential extraction when memory is tight and `k` is small.

Pros and cons by backend
------------------------

| Method | Best for | Pros | Cons |
|---|---|---|---|
| `eigen` | small/medium exact runs | robust reference behavior | can be expensive at scale; very large sparse constraints may trigger truncated eigensolve via `maxeig` |
| `spectra` | larger matrix-free solves | lower memory than dense eigendecomp | iterative behavior can vary by conditioning |
| `randomized` | wide (`p >> n`) low-rank workloads | fast block GEMM/SpMM path | approximation error depends on tuning |
| `deflation` | few components | low memory footprint | can converge slowly; monitor iteration warnings |

Sparse workflow (spectra)
-------------------------

```{r spectra, eval=FALSE}
set.seed(42)
X_sparse <- rsparsematrix(800, 300, density = 0.005)
fit_sp <- genpca(X_sparse, ncomp = 5, method = "spectra",
                 preproc = multivarious::pass(),
                 constraints_remedy = "ridge")
fit_sp$sdev
```

Covariance-only GPCA
--------------------

When you already have the cross-product matrix C = X'MX, `genpca_cov()` avoids working with the full data:

```{r cov}
set.seed(123)
n <- 100; p <- 15
X <- matrix(rnorm(n * p), n, p)
M <- diag(runif(n, 0.8, 1.2))
A <- diag(runif(p, 0.7, 1.3))
C <- t(X) %*% M %*% X
fit_cov <- genpca_cov(C, R = A, ncomp = 5, method = "gmd")
fit_cov$d
```

Out-of-sample projection
------------------------

Fit on training rows, then project held-out observations into the same space:

```{r oos}
set.seed(7)
X <- matrix(rnorm(200 * 30), 200, 30)
fit <- genpca(X[1:150, ], ncomp = 4, preproc = multivarious::center())
scores_test <- multivarious::project(fit, X[151:200, ])
head(scores_test)
```

Performance tips
----------------

Keep data sparse when possible; avoid centering dense copies if you plan to use `spectra`. Use `constraints_remedy = "ridge"` for empirical metrics. For big problems, limit `ncomp` to what you truly need and consider the covariance route if `n` is huge but `p` is moderate.

Where next
----------

See `vignette("gpca-metrics", package = "genpca")` for building metrics and `vignette("gpca-basics", package = "genpca")` for quick-start examples.
