---
title: 'Sparse and Functional PCA with sfpca()'
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: interaction
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts

vignette: |
  %\VignetteIndexEntry{Sparse and Functional PCA with sfpca()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
if (requireNamespace("ragg", quietly = TRUE)) knitr::opts_chunk$set(dev = "ragg_png")
if (requireNamespace("systemfonts", quietly = TRUE)) albersdown::albers_register_fonts()
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
options(rmarkdown.html_vignette.check_title = FALSE)
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  message    = FALSE,
  warning    = FALSE,
  fig.width  = 6,
  fig.height = 4,
  out.width  = "85%"
)
library(genpca)
library(Matrix)

grid_img <- function(z, g, main = "") {
  image(matrix(z, g, g), axes = FALSE, main = main, cex.main = 0.95,
        col = grey.colors(24, start = 0.97, end = 0.12))
  box(col = "grey80")
}
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

`sfpca()` estimates principal components that are **sparse** and **smooth** at
the same time. Ordinary PCA gives you loadings that are dense — every variable
loads on every component — and that spread noise across the whole map. If you
know your variables live somewhere (voxels on a grid, sensors in a room,
wavelengths along a spectrum) and that the real signal is *localised* and
*spatially coherent*, you can ask for both properties directly.

The two requests pull in different directions, which is the point. Sparsity
alone gives you a scatter of isolated survivors; smoothness alone gives you a
blurred map with no zeros anywhere. Together they give compact regions with
soft edges.

## When to reach for it

Use `sfpca()` when both of these hold:

- The loadings should be **mostly zero** — only a minority of variables
  participate in any given component.
- The non-zero part should be **coherent** in some known geometry — neighbours
  should resemble each other.

If you only want structure and not sparsity, `genpca()` with a smoothing
metric is the simpler tool; see `vignette("gpca-metrics")`. If you want
sparsity with no geometry, an ordinary sparse PCA will do. `sfpca()` is for
the case where you want both, and it is worth knowing that it reaches them by
a different mechanism than `genpca()` — see
[Metric form versus constraint form] below, because the two take *opposite*
inputs for the same intent.

## A worked example

Two spatially localised signals on a 16 × 16 grid, each modulated by its own
temporal profile, buried in noise. The spatial patterns are Gaussian bumps
truncated to zero away from their centres, so they are genuinely sparse
(29 of 256 locations) *and* smooth on their support:

```{r simulate}
set.seed(11)
g <- 16; p <- g * g; n <- 64
gr <- expand.grid(r = 1:g, c = 1:g)

# NOTE the orientation: spatial dimensions in ROWS, variables in COLUMNS
spat_cds <- rbind(gr$r, gr$c)
dim(spat_cds)

blob <- function(r0, c0, s = 1.6) {
  z <- exp(-((gr$r - r0)^2 + (gr$c - c0)^2) / (2 * s^2))
  z[z < 0.15] <- 0                       # compact support => sparse
  z / sqrt(sum(z^2))
}
v1 <- blob(5, 5); v2 <- blob(12, 12)

tt <- (0:(n - 1)) / n                    # orthogonal temporal profiles
u1 <- sin(2 * pi * tt); u1 <- u1 / sqrt(sum(u1^2))
u2 <- sin(4 * pi * tt); u2 <- u2 / sqrt(sum(u2^2))

signal <- 30 * tcrossprod(u1, v1) + 20 * tcrossprod(u2, v2)
X <- signal + matrix(rnorm(n * p, sd = 0.25), n, p)

c(true_support = sum(v1 != 0), of = p,
  SNR = round(norm(signal, "F") / norm(X - signal, "F"), 2))
```

The signal-to-noise ratio is close to 1, so this is not a giveaway. Fit two
components:

```{r fit}
fit <- sfpca(X, K = 2, spat_cds = spat_cds)
fit
```

With every penalty left at its default, the loadings come back sparse without
being told how sparse to be:

```{r sparsity}
V <- multivarious::components(fit)
c(nonzero_PC1 = sum(V[, 1] != 0), nonzero_PC2 = sum(V[, 2] != 0), of = p)
```

Against the truth, and against ordinary PCA on the same matrix:

```{r compare}
pc <- prcomp(as.matrix(X), center = TRUE, rank. = 2)

rbind(
  sfpca = c(PC1 = abs(cor(V[, 1], v1)),        PC2 = abs(cor(V[, 2], v2))),
  pca   = c(PC1 = abs(cor(pc$rotation[, 1], v1)), PC2 = abs(cor(pc$rotation[, 2], v2)))
)
```

Correlation alone barely separates them — PCA finds *where* the signal is
perfectly well. The difference is everything else: PCA has to spend all 256
loadings to say it, so the map carries a noise floor everywhere the true
pattern is zero.

```{r recovery-plot, echo = FALSE, fig.cap = "Top row: the true spatial patterns. Middle: sfpca loadings -- compact, and exactly zero away from the bumps. Bottom: ordinary PCA loadings, which recover the location but smear noise across all 256 sites.", fig.width = 6, fig.height = 8.5}
op <- par(mfrow = c(3, 2), mar = c(1, 1, 2.5, 1))
grid_img(v1, g, "True pattern 1");            grid_img(v2, g, "True pattern 2")
grid_img(V[, 1], g, "sfpca loading 1");       grid_img(V[, 2], g, "sfpca loading 2")
grid_img(pc$rotation[, 1], g, "PCA loading 1"); grid_img(pc$rotation[, 2], g, "PCA loading 2")
par(op)
```

The temporal factors are recovered too — those are the `ou` slot, penalised
for roughness along the row index:

```{r temporal}
U <- fit$ou
c(PC1 = abs(cor(U[, 1], u1)), PC2 = abs(cor(U[, 2], u2)))
```

Because the loadings are zero off the bumps, the reconstruction discards noise
that PCA is obliged to keep. Measured against the *noiseless* signal:

```{r reconstruction}
Xhat  <- as.matrix(multivarious::reconstruct(fit))
pchat <- pc$x %*% t(pc$rotation) + matrix(pc$center, n, p, byrow = TRUE)

c(sfpca = norm(signal - Xhat,  "F") / norm(signal, "F"),
  pca   = norm(signal - pchat, "F") / norm(signal, "F"))
```

## What the two penalties do

Each factor carries two penalties, and they are worth separating in your head.
For the column factor `v` those are `lambda_v` (sparsity) and `alpha_v`
(smoothness); `lambda_u` and `alpha_u` do the same for the row factor `u`.
Switching each off in turn shows which is responsible for what:

```{r ablation}
variants <- list(
  "defaults"                  = list(),
  "no sparsity (lambda_v = 0)" = list(lambda_v = 0),
  "no smoothing (alpha_v = 0)" = list(alpha_v = 0),
  "neither"                    = list(lambda_v = 0, alpha_v = 0),
  "heavy sparsity (lambda_v = 3)" = list(lambda_v = 3)
)

t(sapply(variants, function(extra) {
  f <- do.call(sfpca, c(list(X = X, K = 1, spat_cds = spat_cds), extra))
  v <- multivarious::components(f)[, 1]
  c(nonzero = sum(v != 0), cor_with_truth = round(abs(cor(v, v1)), 3))
}))
```

Sparsity is what produces the zeros: drop `lambda_v` and all 256 sites load.
Smoothness does not create zeros — it decides *which* sites survive and keeps
the surviving map coherent. And sparsity can be overdone: at `lambda_v = 3`
the support is cut below the true 29 sites and the recovered pattern degrades.

## Choosing the penalties

Both kinds of penalty have defaults you can usually leave alone.

**Sparsity (`lambda_u`, `lambda_v`).** Left `NULL`, each is chosen per
component by a BIC-style criterion along a regularisation path — `nlambda`
values log-spaced down from a closed-form `lambda_max`, warm-started. The
all-zero solution is a legitimate candidate: a component with no support worth
its degrees of freedom comes back exactly zero with `d = 0`, which is a
feature, not a failure. Fix the value explicitly to bypass the search.

**Smoothness (`alpha_u`, `alpha_v`).** Left `NULL`, each defaults to
$1/\lambda_{\max}(\Omega)$, so the roughest direction of the penalty is
weighted exactly as strongly as the identity term. That makes the default
invariant to how you scaled `Omega` and bounds the condition number of every
inner solve by 2.

What was selected is stored on the fit:

```{r selected}
data.frame(
  component = 1:2,
  lambda_u  = signif(fit$lambda_u, 3), lambda_v = signif(fit$lambda_v, 3),
  alpha_u   = signif(fit$alpha_u, 3),  alpha_v  = signif(fit$alpha_v, 3)
)
```

The penalty *shape* is set by `penalty_u` / `penalty_v`: `"l1"` (the default)
or `"scad"`. SCAD applies less shrinkage to large coefficients, so surviving
loadings keep more of their magnitude, at the price of a non-convex
subproblem.

## Reading the output

`sfpca()` returns a `bi_projector`, so the usual **multivarious** verbs work:
`scores()` for $UD$, `components()` for the sparse loadings $V$, `sdev()` for
$d_k$, and `reconstruct()`. Two things about it differ from `genpca()` and
will bite if you assume otherwise.

The example above is too well behaved to show either, which is itself worth
knowing: its two components were built orthogonal, so they *come out* very
nearly orthogonal and the pitfalls stay hidden. Refit on data whose components
share a temporal profile, and both surface:

```{r correlated-fit}
u2c <- sin(2 * pi * tt + 0.9); u2c <- u2c / sqrt(sum(u2c^2))
round(sum(u1 * u2c), 3)                 # the two profiles now overlap

set.seed(11)
Xc  <- 30 * tcrossprod(u1, v1) + 20 * tcrossprod(u2c, v2) +
       matrix(rnorm(n * p, sd = 0.25), n, p)
fc  <- sfpca(Xc, K = 2, spat_cds = spat_cds)
Uc  <- fc$ou; Vc <- multivarious::components(fc)
```

**The factors are not orthogonal.** Each rank-1 term comes from its own
constraint-form subproblem rather than a joint SVD. Columns are unit-norm, but
$U^{\top}U \ne I$ and $V^{\top}V \ne I$ in general:

```{r orthogonality}
round(crossprod(Uc), 3)     # would be the identity for a joint SVD
round(crossprod(Vc), 3)
```

**`sdev()` is not the singular values of `X`.** It is the covariance each
component captures, $d_k = u_k^{\top} X_k v_k$, where $X_k$ is the matrix
*after* the previous components have been deflated out. Only the first
component is measured against the original data:

```{r sdev-meaning}
Xm   <- as.matrix(Xc)
dc   <- multivarious::sdev(fc)
defl <- Xm - dc[1] * tcrossprod(Uc[, 1], Vc[, 1])

c(sdev_2      = dc[2],
  u2_X_v2     = as.numeric(t(Uc[, 2]) %*% Xm   %*% Vc[, 2]),   # does NOT match
  u2_Xdefl_v2 = as.numeric(t(Uc[, 2]) %*% defl %*% Vc[, 2]))   # matches
```

The gap is small here but it is not noise, and it grows with how much the
components share. Treat `sdev()` as "covariance captured by this component
given the previous ones", never as a singular value of `X`.

Because $V$ is not orthogonal, `reconstruct()` multiplies the stored factors
directly as $UDV^{\top}$ rather than going through pseudo-inverse identities,
which would not reproduce the fitted model.

## Metric form versus constraint form

This is the trap when moving between `sfpca()` and `genpca()`: **they take
opposite inputs for the same intent.**

In `genpca()`, the structure matrix $A$ is a *metric*, and a metric amplifies
its own dominant eigendirections — the loadings are $AV$. To get smooth
loadings you pass a **smoother** (a kernel, an adjacency matrix,
$(I + \alpha\Omega)^{-1}$).

In `sfpca()`, the same information enters as a *constraint*,
$v^{\top}(I + \alpha\Omega)v \le 1$, which charges rough $v$ against a fixed
budget. So you supply the **roughness operator** directly, and a larger
`alpha_v` means a smoother result.

| | `genpca()` | `sfpca()` |
|:--|:--|:--|
| Structure enters as | metric $A$ | constraint $v^{\top}(I+\alpha\Omega)v \le 1$ |
| For smooth loadings, supply | a smoother (kernel, $(I+\alpha\Omega)^{-1}$) | a roughness operator ($\Omega$, a Laplacian) |
| Turning the knob up | amplifies $A$'s top directions | smooths more |

The same Laplacian therefore *smooths* in `sfpca()` and *roughens* in
`genpca()`. `vignette("gpca-metrics")` works through the metric-side version
of this in detail.

## Practical notes

**`spat_cds` is dimensions × variables.** Rows are spatial axes, columns are
variables — the transpose of the usual data-frame layout. Getting it backwards
does not give a clean message; on this example it fails with a warning about
having only 2 locations, then a non-conformable-dimensions error:

```{r orientation-trap, error = TRUE, warning = TRUE}
sfpca(X, K = 1, spat_cds = t(spat_cds))
```

Neither message points at the real problem, so check `dim(spat_cds)` first if
you hit it.

**The column penalty is built for you.** `Omega_v` is constructed internally
from `spat_cds` via a `knn` nearest-neighbour graph (default
`min(6, ncol(X) - 1)`); there is no `Omega_v` argument. `Omega_u` *can* be
supplied, and defaults to a second-difference operator — which assumes the
rows are ordered, as with a time series. If your rows are unordered samples,
pass `alpha_u = 0` rather than smoothing along a meaningless axis.

**Components are extracted by deflation,** so cost grows linearly in `K` and
later components are fit to residuals. Ask for the number you intend to
interpret.

## Where next

`vignette("gpca-metrics")` covers the metric-side treatment of the same
structural ideas, including how to build kernels, Laplacians and graph
penalties. `vignette("structured-noise")` discusses choosing between them when
several kinds of structure are present at once.

## Reference

Allen, G. I., & Weylandt, M. (2019). Sparse and functional principal
components analysis. In *2019 IEEE Data Science Workshop (DSW)* (pp. 11--16).
doi:[10.1109/DSW.2019.8755778](https://doi.org/10.1109/DSW.2019.8755778).
Also available as arXiv:[1309.2895](https://arxiv.org/abs/1309.2895), first
posted in 2013 and revised through 2019 — the preprint and the DSW paper are
the same work, which is why the literature cites both years.
