---
title: Modelling Structured Noise
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: interaction
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts

vignette: |
  %\VignetteIndexEntry{Modelling Structured Noise}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
if (requireNamespace("ragg", quietly = TRUE)) knitr::opts_chunk$set(dev = "ragg_png")
if (requireNamespace("systemfonts", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) albersdown::albers_register_fonts()
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
options(rmarkdown.html_vignette.check_title = FALSE)
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  message    = FALSE,
  warning    = FALSE,
  fig.width  = 6,
  fig.height = 4,
  out.width  = "85%"
)
library(genpca)
library(Matrix)
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```


## The problem

Real data rarely offers a clean choice. An fMRI run has smooth signal
(distributed networks), smooth noise (drift, vascular and physiological
fluctuation), rough noise (thermal), and sometimes rough signal (focal
activation, tissue boundaries) — all at once. The obvious question is which
metric to reach for, and the honest answer is that the choice is
better-defined than it looks in some cases and impossible in others.

This vignette works out which is which. Every number below is produced by the
code shown.

## A metric is a spectral filter

Build a graph over your variables — voxels adjacent in space, time points
adjacent in a series — and take its Laplacian $L$. The eigenvectors of $L$
are a Fourier basis for that graph: small eigenvalues are smooth patterns,
large eigenvalues are rough ones. Then *any* metric of the form

$$A = f(L) = \sum_j f(\lambda_j)\,\phi_j\phi_j^{\top}$$

is a filter, and $f$ is its transfer function. That single observation covers
every recipe in this package.

```{r graph-setup}
g <- 12; p <- g * g; n <- 150
gi <- expand.grid(r = 1:g, c = 1:g)
W <- matrix(0, p, p)
for (i in 1:p) for (j in 1:p)
  if (i < j && abs(gi$r[i] - gi$r[j]) + abs(gi$c[i] - gi$c[j]) == 1) {
    W[i, j] <- 1; W[j, i] <- 1
  }
L  <- diag(rowSums(W)) - W
eL <- eigen(L, symmetric = TRUE)
Q  <- eL$vectors
lam <- pmax(eL$values, 0)

# build a metric from a transfer function of the Laplacian
spec <- function(f) Q %*% (f(lam) * t(Q))
```

Note that `eigen()` returns eigenvalues in *decreasing* order, so `Q[, 1]` is
the roughest graph mode and `Q[, p]` the smoothest. Getting this backwards is
an easy way to convince yourself of something false.

```{r profiles, echo = FALSE, fig.cap = "Four transfer functions on the same graph. Everything to the left of the crossing is smooth structure; everything to the right is rough. A metric multiplies the decomposition's attention by this curve.", fig.height = 3.6}
ls <- seq(0, max(lam), length.out = 200)
prof <- cbind(smoother  = 1 / (1 + 6 * ls),
              precision = 1 / (0.02 + 1 / (1 + 6 * ls)),
              unbounded = 1 + 6 * ls,
              bandpass  = exp(-((ls - 3.5)^2) / 1.5))
prof <- sweep(prof, 2, apply(prof, 2, max), "/")
matplot(ls, prof, type = "l", lty = 1, lwd = 2,
        col = c("steelblue", "tomato", "grey30", "darkolivegreen"),
        xlab = expression(paste("Laplacian eigenvalue  ", lambda,
                                "   (smooth ", symbol("\256"), " rough)")),
        ylab = "relative weight f(lambda)")
legend("right", c("smoother", "bounded precision", "unbounded I + aL", "band-pass"),
       col = c("steelblue", "tomato", "grey30", "darkolivegreen"),
       lty = 1, lwd = 2, bty = "n", cex = 0.8)
```

So the practical vocabulary is not "adjacency versus Laplacian" — those are
just two points on this continuum. You are choosing a curve.

### Building these matrices in practice

You rarely have to assemble a graph by hand. The
[adjoin](https://cran.r-project.org/package=adjoin) package constructs
weighting matrices from coordinates or from the data itself, and supplies
both orientations directly: `spatial_adjacency()`, `spatial_smoother()`,
`heat_kernel()` and `graph_weights()` on the smoother side;
`spatial_laplacian()` and `temporal_laplacian()` on the precision side, with
`temporal_adjacency()` for the time margin.

```{r adjoin, eval = requireNamespace("adjoin", quietly = TRUE)}
cds  <- as.matrix(expand.grid(x = 1:8, y = 1:8))
Aadj <- adjoin::spatial_adjacency(cds, nnk = 8, weight_mode = "heat", sigma = 1.5)
Alap <- adjoin::spatial_laplacian(cds, nnk = 8, weight_mode = "heat", sigma = 1.5)

# check the two traps before using either as a metric
range(eigen(as.matrix(Aadj), symmetric = TRUE, only.values = TRUE)$values)
range(eigen(as.matrix(Alap), symmetric = TRUE, only.values = TRUE)$values)
```

Both traps from the previous section show up in real output. The adjacency
comes back **indefinite** — its smallest eigenvalue is negative, so it is not
a metric until you shift it (`Aadj + c * Diagonal(n)`) or let
`constraints_remedy` repair it. The Laplacian comes back **PSD but exactly
singular**, its null vector being the spatially constant pattern; `genpca()`
accepts that via a pseudo-inverse, but `Alap + eps * Diagonal(n)` is usually
what you want. Neither is a defect in `adjoin` — an adjacency matrix simply
is not positive definite, and that is a property of graphs, not of the
software.

With those repairs, the two point in the directions the table above predicts:
used as a column metric, the adjacency produces markedly smoother loadings
than plain PCA and the Laplacian markedly rougher ones.

## What a metric can and cannot separate

Here is the part that decides everything. A spatial metric reweights by
*spatial frequency*, and by nothing else. It can therefore separate signal
from noise exactly when the two occupy different parts of the spectrum.

We plant a signal of known spatial character in noise of known spatial
character, and measure how well the leading component recovers it.

```{r twobytwo}
nrm  <- function(v) v / sqrt(sum(v^2))
# how much of the planted pattern is captured by the fitted subspace?
recov <- function(V, truth) {
  V <- qr.Q(qr(as.matrix(V)))
  sqrt(sum((t(V) %*% truth)^2)) / sqrt(sum(truth^2))
}

smooth_pat <- nrm(exp(-((gi$r - 4)^2 + (gi$c - 4)^2) / 6))     # a blob
fine_pat   <- nrm(Q[, which.min(abs(lam - median(lam)))])      # a mid-frequency mode

smooth_noise <- function(n) {
  Z <- matrix(rnorm(n * p), n, p) %*% spec(function(l) sqrt(1 / (1 + 4 * l)))
  Z / sqrt(mean(Z^2))
}
fine_noise <- function(n) {
  Z <- matrix(rnorm(n * p), n, p) %*% spec(function(l) sqrt((l + .5) / max(lam)))
  Z / sqrt(mean(Z^2))
}

A_smoother  <- spec(function(l) 1 / (1 + 6 * l))
A_precision <- spec(function(l) 1 / (0.02 + 1 / (1 + 6 * l)))

set.seed(909)
reps <- 8
grid <- expand.grid(signal = c("smooth", "fine"), noise = c("smooth", "fine"),
                    stringsAsFactors = FALSE)
out <- t(apply(grid, 1, function(row) {
  pat <- if (row[["signal"]] == "smooth") smooth_pat else fine_pat
  acc <- c(0, 0, 0)
  for (r in seq_len(reps)) {
    E <- if (row[["noise"]] == "smooth") smooth_noise(n) else fine_noise(n)
    X <- scale(matrix(rnorm(n), n, 1) %*% t(pat) * 1.1 + E, scale = FALSE)
    for (k in 1:3) {
      A <- list(diag(p), A_smoother, A_precision)[[k]]
      fit <- genpca(X, A = A, ncomp = 1, preproc = multivarious::pass())
      acc[k] <- acc[k] + recov(fit$ov, pat) / reps
    }
  }
  acc
}))
dimnames(out) <- list(paste(grid$signal, "signal /", grid$noise, "noise"),
                      c("identity", "smoother", "precision"))
round(out, 3)
```

Read the rows. When signal and noise sit in **different** bands, the matching
metric is transformative — better than tenfold in both mismatched rows: the
smoother rescues a smooth signal from fine noise, and the precision rescues a
fine signal from smooth noise, in each case lifting recovery from near-zero
to roughly 0.6–0.7. Note also that in those rows the *wrong* metric is no
better than identity and sometimes worse, so the gain is genuinely about
matching the filter to the gap between signal and noise.

When signal and noise sit in the **same** band, the picture changes
completely: all three columns land in the same range, with no metric
delivering anything like the off-diagonal gain. There is no filter that
separates two things occupying the same frequencies.

|  | smooth noise | fine noise |
|:--|:--|:--|
| **smooth signal** | no metric helps | smoother |
| **fine signal** | precision | no metric helps |

This is the answer to "my data has all four at once." Decompose the problem
by band, not by wish. The components of the noise that differ spectrally from
your signal can be suppressed by $A$; the components that overlap it cannot,
and pretending otherwise just costs you signal.

## Whiten by the noise, not by the signal

The rule for choosing $f$ follows from the model rather than from taste.
Under separable noise, $A = \Sigma_{\text{col}}^{-1}$ makes GPCA the
maximum-likelihood low-rank fit, so the metric is determined by the
**noise**, whatever the signal happens to look like.

For a realistic two-component noise model — smooth physiological fluctuation
plus broadband thermal noise —

$$\Sigma_{\text{col}} = \sigma_s^2\,K + \sigma_r^2 I
\qquad\Longrightarrow\qquad
A = \left(\sigma_s^2\,K + \sigma_r^2 I\right)^{-1},$$

which is the "bounded precision" curve in the first figure. It suppresses the
smooth band where the physiological noise lives and then *flattens out* at
$1/\sigma_r^2$ instead of growing without limit. Two knobs, both with physical
meaning, both estimable from resting or baseline data.

Estimate them from data that does not contain your effect — a baseline run,
or the residuals after removing the design — rather than from the data you
are about to decompose.

## Bounded beats unbounded

The reason to insist on that flattening: an unbounded metric such as
$I + \alpha L$ keeps increasing with $\lambda$, so it places its largest
weight on the very roughest directions. That is precisely where thermal noise
lives and where signal generally does not.

```{r bounded}
A_unbounded <- spec(function(l) 1 + 6 * l)

set.seed(78)
acc <- c(0, 0, 0); reps <- 8
for (r in seq_len(reps)) {
  X <- scale(matrix(rnorm(n), n, 1) %*% t(fine_pat) * 1.1 +
             smooth_noise(n) * 0.8 +
             matrix(rnorm(n * p), n, p) * 0.8, scale = FALSE)   # + thermal
  for (k in 1:3) {
    A <- list(diag(p), A_precision, A_unbounded)[[k]]
    fit <- genpca(X, A = A, ncomp = 1, preproc = multivarious::pass())
    acc[k] <- acc[k] + recov(fit$ov, fine_pat) / reps
  }
}
setNames(round(acc, 3), c("identity", "bounded precision", "unbounded I + 6L"))
```

Both precisions beat plain PCA by a wide margin, and the bounded one comes
out ahead: once broadband thermal noise is in the mixture, spending unlimited
weight at the top of the spectrum buys noise rather than signal. The margin
here is modest because the planted signal is only mid-frequency; it widens as
the thermal component grows or as the signal sits further from the top of the
spectrum. Prefer the saturating form — its worst case is bounded, and the
unbounded one has no worst case at all.

## Use both margins

The cases the table above marks "no metric helps" are not hopeless — they are
just not solvable *in space*. A smooth task response and a smooth drift are
indistinguishable by spatial frequency, but they are utterly different in
time. That is what the row metric is for, and it is why `M` and `A` are
separate arguments rather than one blended constraint.

Here the noise is temporally autocorrelated, the signal is task-locked at a
frequency where that noise has little power, and the row metric is the AR(1)
precision — the same prewhitening used in a standard fMRI GLM.

```{r rowmetric}
rho   <- 0.85
Sig_t <- outer(0:(n - 1), 0:(n - 1), function(i, j) rho^abs(i - j))
M_ar  <- solve(Sig_t + 1e-6 * diag(n))
task  <- scale(sin(2 * pi * (1:n) / 7))

set.seed(303)
acc <- c(0, 0); reps <- 8
for (r in seq_len(reps)) {
  E <- t(chol(Sig_t)) %*% matrix(rnorm(n * p), n, p)      # AR(1) in time
  X <- scale(task %*% t(smooth_pat) + E, scale = FALSE)
  acc[1] <- acc[1] + recov(genpca(X, ncomp = 1,
                                  preproc = multivarious::pass())$ov, smooth_pat) / reps
  acc[2] <- acc[2] + recov(genpca(X, M = M_ar, ncomp = 1,
                                  preproc = multivarious::pass())$ov, smooth_pat) / reps
}
setNames(round(acc, 3), c("no row metric", "AR(1) precision M"))
```

Roughly a tenfold improvement on the same data, with the column metric left
as identity throughout. The temporal structure did the work that no spatial
metric could.

The practical consequence for imaging: put temporal nuisances on `M` (AR
prewhitening, down-weighting motion-corrupted frames) and spatial nuisances
on `A`, and regress out what is better removed by design — drift terms,
physiological regressors — before decomposing at all.

## When one metric is not enough

A single $A$ applies **one** filter to **every** component. If your data
contains a smooth network component *and* a focal artifact, no single choice
serves both. Two escape routes:

- `sfpca()` selects a sparsity penalty per component by BIC, so different
  components can have different spatial extent. Note its structure argument
  is a *constraint*, $v^{\top}(I + \alpha\Omega)v \le 1$, so you pass the
  roughness operator directly and larger `alpha_v` means smoother — the
  opposite convention to a metric.
- `gpca_mle()` and `mnpca_mrl()` learn `M` and `A` from the data by penalized
  maximum likelihood instead of asking you to specify them, the latter with
  sparse precision matrices.

## Caveats worth carrying

**Separability is an approximation.** The whole framework assumes noise
covariance factorizes as $\Sigma_{\text{space}} \otimes \Sigma_{\text{time}}$.
That is fair for thermal noise and for correlation induced by smoothing. It
is not fair for cardiac and respiratory noise, which is spatially localized
near vessels and ventricles *and* temporally periodic; that structure
violates the Kronecker form rather than being absorbed by it.

**Whitening does not create signal.** Equalizing the noise floor changes
which directions dominate the decomposition; it does not change the
signal-to-noise ratio within any direction. A signal below the noise in its
own band stays below it.

**Validate on your own data.** Every number here comes from a simulation
whose ground truth we chose. Sweep the strength of your metric, look at the
loadings, and confirm the components move the way you expect before
believing a result that depends on the metric.

## Where next

- `vignette("gpca-metrics")` — concrete recipes and the smoother/precision orientation rule.
- `vignette("genpca")` — the decomposition itself, and what the metrics mean geometrically.
- `vignette("gpca-scale")` — backends for when these metrics get large.
